Formalization of Forcing in Isabelle/ZF

### 2020-06-17 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
highliting in slides, script almost complete
slides: minor. script: ZF-Constructible

\begin{document}
%\nobibliography*

@@

\begin{frame}
  \frametitle{Why Isabelle/ZF?}
  \begin{block}{Pros}
    \begin{itemize}
    \item Most advanced set theory formalized (around 2017).
      % \item Closer to the mathematical practice of forcing.
    \item Structured proof language Isar
      \citep{DBLP:conf/tphol/Wenzel99}.
    \item Low in consistency strength.
    \end{itemize}
  \end{block}
  %
  \pause
  %
  \begin{block}{Cons}
    \begin{itemize}
    \item A fraction of automation of Isabelle (\texttt{sledgehammer},
      etc).
    \item ``Untyped'', and too weak a metatheory.
    \end{itemize}
  \end{block}
\end{frame}

@@

%% \begin{frame}
%%   \frametitle{A word on consistency}
%% \end{frame}

@@

\begin{frame}
  \frametitle{Some details}
  \begin{enumerate}%\setcounter{enumi}{1}
  \item We adapted \texttt{ZF-Constructible} to obtain
    sharper absoluteness results.
  \end{enumerate}

  This library contains a major development of relativization and
  absoluteness for \emph{classes} $C :: {}$ \isatt{i} $\Rightarrow$
  \isatt{o}, including the definition of the set \isatt{formula} ::
  \isatt{i} of internal formulas and model theoretic satisfaction. 
  \pause
  \begin{block}{Relativization and synthesis discipline}
    \begin{enumerate}[I]
    \item<+-> $p=\{x,y\} :: {}$\isatt{i}
      \hfill
      (original term).
    \item<+-> \isatt{upair}$(C,x,y,p) :: {}$\isatt{o}
      \hfill
      (relativization, fully
      relational). 
    \item<+-> \isatt{upair{\uscore}fm}$(0,1,2) :: {}$\isatt{i} 
      \hfill
      (synthesized member of \isatt{formula}).
    \end{enumerate}
  \end{block}
  \begin{alertblock}{}<+->
    Around 40 absoluteness/closure lemmas now hold using weaker
    hypotheses on the class $C$ (most of them, just that $C$ is
    transitive and nonempty).
  \end{alertblock}
\end{frame}

\begin{frame}{Some details}
  \begin{enumerate}\setcounter{enumi}{1}
  \item We formalized the formula transformer $\forceisa$ and
    hence the forcing relation $\forces$, and proved the Fundamental
    Theorems.
  \end{enumerate}
  %
  \pause
  %
  \begin{block}{$\forceisa$ is defined by recursion on  formulas}
    This is the main reason we work with the set of internalized
    formulas, and that we require legit first-order expressions for
    the axiom schemes (Separation and Replacement).

@@

\begin{frame}{Some details}
  \begin{enumerate}\setcounter{enumi}{2}
  \item For any ctm $M$ of $\ZF$ and $M$-generic $G$, $M[G] \models
    \ZF$ (respectively, adding $\AC$).
  \end{enumerate}
  \pause
  Actually, the modularity of the theory of forcing allowed us to do
  this \alert{before} we write down the definition of $\forceisa$.
  \pause

@@

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% \section{}
%% %-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%-%
%% \subsection{}
%% 
%% \begin{frame}
%%   \frametitle{}
%% \end{frame}

@@

\begin{frame}
  \frametitle{Extra: Locale structure involving set models} 
             {  
               \renewcommand{\arraystretch}{1.5}               \begin{tabular}{rcl}
                 \texttt{forcing{\uscore}notion} & = & preorder $\PP$ with top. \\
                 %    \texttt{M{\uscore}ZF} & = & set model of the $\ZF$ axioms \\ 
                 \texttt{M{\uscore}ZF{\uscore}trans} & = & set model $M$ of the $\ZF$
                 axioms \alert{+}  $M$ transitive \\ 
                 \texttt{M{\uscore}ctm} & = &  \texttt{M{\uscore}ZF{\uscore}trans} \alert{+}
                 $M$ countable \\
                 \texttt{forcing{\uscore}data} & =  & \texttt{M{\uscore}ctm} \alert{+}
                 \texttt{forcing{\uscore}notion} $\PP\in M$\\
                 \texttt{separative{\uscore}notion} & = &
                 \texttt{forcing{\uscore}notion} \alert{+} $\PP$ separative \\
                 \texttt{M{\uscore}ctm{\uscore}separative} & = &
                 \texttt{forcing{\uscore}data} \alert{+}
                 \texttt{separative{\uscore}notion} \\
                 \texttt{G{\uscore}generic} & = & \texttt{forcing{\uscore}data} \alert{+} $G$ is $M$-generic
               \end{tabular} 
             }
\end{frame}

@@

%% - Más bola a la metateoría (ZF).
%% 
%% * es weak ---> falta de inducción/recursión, y el objeto
%% central (forces) se define así.
%% * too strong ---> not possible to have finitary consistency proofs
%%

>>>>

\begin{document}

@@

\begin{frame}
  \frametitle{Why Isabelle/ZF?}
  \begin{block}{Pros}
    \begin{itemize}
    \item Most advanced set theory formalized (around 2017).
    \item Structured proof language Isar
      \citep{DBLP:conf/tphol/Wenzel99}.
    \item Comparatively low in consistency strength.
    \end{itemize}
  \end{block}
  %
  \pause
  %
  \begin{block}{Cons}
    \begin{itemize}
    \item A fraction of automation of Isabelle (\texttt{sledgehammer},
      etc).
    \item ``Untyped'', and too weak a metatheory.
    \end{itemize}
  \end{block}
\end{frame}

@@

@@

\begin{frame}
  \frametitle{Some details}
  \begin{alertblock}{}
    \begin{enumerate}%\setcounter{enumi}{1}
    \item We adapted \texttt{ZF-Constructible} to obtain
      sharper absoluteness results.
    \end{enumerate}
  \end{alertblock}

  This library originally contains a major development of relativization and
  absoluteness for \emph{classes} $C :: {}$ \isatt{i} $\Rightarrow$
  \isatt{o}, including the definition of the set \isatt{formula} ::
  \isatt{i} of internal formulas and model theoretic satisfaction. 
  \pause
  \begin{block}{Relativization and synthesis discipline}
    \begin{enumerate}[I]
    \item<+-> $p=\{x,y\} :: {}$\isatt{i}
      \hfill
      (original term).
    \item<+-> \isatt{upair}$(C,x,y,p) :: {}$\isatt{o}
      \hfill
      (relativization, fully
      relational). 
    \item<+-> \isatt{upair{\uscore}fm}$(0,1,2) :: {}$\isatt{i} 
      \hfill
      (synthesized member of \isatt{formula}).
    \end{enumerate}
  \end{block}
  \begin{alertblock}{}<+->
    Around 40 absoluteness/closure lemmas now hold using weaker
    hypotheses on the class $C$ (most of them, just that $C$ is
    transitive and nonempty).
  \end{alertblock}
\end{frame}

\begin{frame}{Some details}
  \begin{alertblock}{}
    \begin{enumerate}\setcounter{enumi}{1}
    \item We formalized the formula transformer $\forceisa$ and
      hence the forcing relation $\forces$, and proved the Fundamental
      Theorems.
    \end{enumerate}
  \end{alertblock}
  %
  \pause
  %
  \begin{block}{$\;\forceisa$ is defined by recursion on  formulas}
    This is the main reason we work with the set of internalized
    formulas, and that we require legit first-order expressions for
    the axiom schemes (Separation and Replacement).

@@

\begin{frame}{Some details}
  \begin{alertblock}{}
    \begin{enumerate}\setcounter{enumi}{2}
    \item For any ctm $M$ of $\ZF$ and $M$-generic $G$, $M[G] \models
      \ZF$ (respectively, adding $\AC$).
    \end{enumerate}
  \end{alertblock}
  \pause
  Actually, the modularity of the theory of forcing allowed us to do
  this \alert{before} we write down the definition of $\forceisa$.
  \pause

@@

@@

\begin{frame}
  \frametitle{Extra: Locale structure involving set models} 
             {  
               \renewcommand{\arraystretch}{1.5}               \begin{tabular}{rcl}
                 \texttt{forcing{\uscore}notion} & = & preorder $\PP$ with top. \\
                 \texttt{M{\uscore}ZF{\uscore}trans} & = & set model $M$ of the $\ZF$
                 axioms \alert{+}  $M$ transitive \\ 
                 \texttt{M{\uscore}ctm} & = &  \texttt{M{\uscore}ZF{\uscore}trans} \alert{+}
                 $M$ countable \\
                 \texttt{forcing{\uscore}data} & =  & \texttt{M{\uscore}ctm} \alert{+}
                 \texttt{forcing{\uscore}notion} $\PP\in M$\\
                 \texttt{separative{\uscore}notion} & = &
                 \texttt{forcing{\uscore}notion} \alert{+} $\PP$ separative \\
                 \texttt{M{\uscore}ctm{\uscore}separative} & = &
                 \texttt{forcing{\uscore}data} \alert{+}
                 \texttt{separative{\uscore}notion} \\
                 \texttt{G{\uscore}generic} & = & \texttt{forcing{\uscore}data} \alert{+} $G$ is $M$-generic
               \end{tabular} 
             }
\end{frame}

@@
